Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-06-02 VegOzone.tex
Fine tuning

In Section~\ref{sec:dam}, we give a brief summary of the key components and meteorological input data of the $\mathrm{DO_3SE}$ model that we use to assess the ozone risk. Subsequently, we propose a local climate derived temperature response function and explore the impact of the stomata aperture parameterization at low light intensities. Start and end of the growing season are inferred from satellite observed gross primary productivity (GPP). We explore how risk might be over- or underestimated using Boreal parameterization (Section~\ref{sec:results}). In Section~\ref{subsec:res_local}, we evaluate the local climate derived parameterizations by means of average relative stomatal conductance. For evaluating the $\mathrm{DO_3SE}$ model results in a climate perspective, we assess the climate anomalies of weather conditions and ozone concentrations in the 2018/19 GS at our research site (Section~\ref{sec:stats}). In Section~\ref{subsec:do3se_results}, we assess the difference in predicted risk for ozone-induced damage for the local climate derived parameterizations compared to MM parameters. In Section~\ref{sec:conc}, we discuss our results concerning prevailing conditions and give an outlook on how future assessments of pollution risk in the subarctic could be improved.

%%
\section{Methods}
\label{sec:dam}
In the following, we briefly describe the $\mathrm{DO_3SE}$ model used for assessing the ozone risk, its key components, input data, and gap filling procedure (Section~\ref{sec:do3se}). To study the impact of model parameters, we propose a method to define a parameterization for a subarctic climate based on existing Boreal parameterizations and local climate conditions, study the impact of stomata aperture at low light intensities, and estimate the start and end of the growing season from satellite derived GPP (Section~\ref{sec:localization}).

@@

The aperture of stomata referred to as $g_\mathrm{sto}$ is governed by different species-specific factors and responds to key environmental variables. In the $\mathrm{DO_3SE}$ model, $g_\mathrm{sto}$ is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g_\text{sto} = g_\text{max} \cdot f_\text{phen} \cdot f_\text{light} \cdot \max{\left\{f_\text{min}, f_\mathrm{T} \cdot f_\mathrm{VPD} \cdot f_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is modified by factors that vary within a range of $0-1$. These factors are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_\mathrm{T}$), water vapor pressure deficit ($f_\mathrm{VPD}$), and soil water potential ($f_\text{SWP}$). All factors differ with plant functional type (PFT).

The leaf phenology factor ($f_\mathrm{phen}$) is a stepwise defined linear function that allows for an increase of stomatal conductance in spring (start of growing season) and a decrease in fall (end of growing season).

@@

$\Phi_\mathrm{sto}$ for a sunlit upper canopy leaf (vegetation height $h$) is then calculated under the assumption that the concentration of \chem{O_3} at the top of the canopy $h$ represents a reasonable estimate of the \chem{O_3} concentration at the upper surface of the laminar layer:
%
\begin{equation}
  \Phi_\mathrm{sto}(t) = \chem{[O_3]}_h(t) \cdot \frac{g_\mathrm{sto}}{g_\mathrm{sto}+g_\mathrm{ext}} \cdot \frac{1}{r_b+r_c}.
  \label{eq:flux_stomata}
\end{equation}

$\mathrm{POD_y}$ is then computed according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}(t)-y)\cdot \mathrm{d}t},
  \label{eq:pod}
\end{equation}
%
with an hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}(t)$ (Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}^{-1}\,s^{-1}}. Herein PLA stands for Projected~Leaf~Area. The timestep $\mathrm{d}t$ is usually 1~hour. The flux threshold~$\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). Finally, $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

@@

To fill gaps in the ozone data, we applied the methodology described in \citet{ACP:Falk2021} that is based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology (multiannual norm) \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. \citet{ACP:Falk2021} computed a generalized climatology for northern Fennoscandia from long-term records of different close by sites and determined the station with the highest correlation coefficient compared to the site with the gap. For this site, the anomalies with respect to the generalized climatology were calculated for the gap period. These anomalies were than added to the climatology of the site with the gap and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details and a comparison with regional air quality models refer to \citet{ACP:Falk2021}.

@@

For our study focuses on ozone impact on subarctic vegetation, we chose Svanhovd in northern Norway ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$) as site. Reasons for our choice were: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution monitoring (\chem{NO_x}, \chem{SO_x}, particular matter (PM), radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018/19 growing season.

All essential climate variables for the 2018/19 growing season were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The \chem{O_3} monitoring data have been added to a database for air-quality monitoring called EBAS operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration, drifts in the observed data occurred imposing false trends which are small but in the same order of magnitude as the expected trends in tropospheric background ozone. This does not concern the ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible injuries on ozone sensitive plants, following \citet{GCB:Mills2011}, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed visible ozone injuries on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible injuries were found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). Note that the extent of visible injuries is, in general, not equivalent to a quantitative reduction of above/below ground biomass of a species.

@@

We determined the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from Boreal paramterizations \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. The subarctic local climate derived parameterization for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on the MM's silver birch, and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of MM parameters, consult Supplement~Table~S1.

@@

Initial simulations with the Central European (Boreal is not available) MM parameterizations for perennial grassland revealed an unexpectedly low leaf level stomatal conductance ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occurred only during an extended warm period in late July. In ecological terms, this means that there would have been almost no growth of grass in the summer of 2019 -- a model prediction that is falsified by reality. We identified $f_\mathrm{T}$ as abnormally low throughout the season, being the limiting factor of stomatal conductance in perennial grassland. This suggested that deriving a local climate parameterization is necessary for appropriate modeling.

To derive a subarctic parameterization for the temperature response function $f_\mathrm{T}$, we divided the temperature dataset into two subsets: 1992--2000 (near past) and 2011--2019 (present). From these two subsets, we selected only dates within the GS (May--September) and plotted the probability density functions (PDFs) as histograms with $1\,\unit{^\circ C}$ binning (Fig.~\ref{fig:f_temp_grassland}a). A comparison between the two PDFs indicates that during the past decade the temperature distribution has shifted towards higher temperatures reducing the number of days below $5\,\unit{^\circ C}$. It is unclear how fast the species composition or physiology of perennial grassland responds to these changes. Thus, the predominant acclimation at present is not known. Further, we overlaid these histograms with the MM $f_\mathrm{T}$. A low overlapping area suggests a low agreement with local climate conditions.

We defined local climate derived parameterizations for two hypothetical species (\emph{subarctic} and \emph{cold}) by increasing the overlapping area between $f_\mathrm{T}$ and the temperature PDF during the GS. We constructed \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures than the MM species. This was accomplished by moving $T_\mathrm{min}$ and $T_\mathrm{opt}$ towards cooler temperatures but $T_\mathrm{max}$ was kept at its MM value. In the same way, \emph{subarctic} was constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shifted both $T_\mathrm{min}$ and $T_\mathrm{max}$ to colder temperatures and chose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. The most extreme of the two local climate derived parameterizations compared to the MM is the \emph{subarctic}. This methodology is analogous to the optimal temperature acclimation experimentally found for photosynthesis \citep{PR:Way2013}.

For localizing $f_\mathrm{light}$, we presumed that the aperture of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, adjusted the modelled light sensitivity of stomata such that the PPFD value needed to reach $50\,\unit{\%}$ opening is varied by $\pm 20\,\unit{\%}$. For this, we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT $k$ analytically
%
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
%
First, we solved Eq.~(\ref{eq:inverse_function}) for the MM default value of $\alpha_\mathrm{MM}^k$ at $0.5$ ($50\,\%$ opening)
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and defined a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By solving Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$.

In Fig.~\ref{fig:f_temp_grassland}b, we show the PPFD PDF for near past and present together with the $f_\mathrm{light}$ for perennial grassland. As no Boreal parameterization is available for perennial grassland, the Central European MM parameterization is displayed as solid line.

@@

Similarly, we defined local climate derived $f_\mathrm{T}$ and localized $f_\mathrm{light}$ for coniferous and deciduous trees. The resulting response functions are shown in Appendix~Figs.~\ref{fig:f_temp_spruce} and \ref{fig:f_temp_birch}, respectively.
Coniferous trees are known to start photosynthesis as soon as the leaf temperature reaches a certain threshold \citep{TB:Kolari2007}. Thus, coniferous trees are already active at low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as day of the year~(\unit{doy})~100 but cease activity in less favorable conditions \citep{TB:Kolari2007, TP:Wallin2013}. 
To determine the local climate derived temperature parameterization for coniferous trees, we used the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007}. From these, we estimated an optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). For simplicity, the same interval was assumed for deciduous trees.

@@

\begin{table}[t]
  \caption{Local climate derived light parameterizations. MM refers to mapping manual \citep{GCB:Mills2011,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_light}
  \begin{tabular}{llcr}
    \tophline
    Species & type & $\alpha$ & $\gamma(0.5)$\\
    \middlehline
    \multirow{3}{*}{Deciduous tree} & MM & 0.004 & 165.035\\
    & PPFD0.8 & 0.005 & 132.028\\
    & PPFD1.2 & 0.003 & 198.042\\
    \middlehline
    \multirow{3}{*}{Coniferous tree} & MM & 0.006 & 115.525\\
    & PPFD0.8 & 0.008 & 92.420\\
    & PPFD1.2 & 0.005 & 138.629\\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 0.011 & 63.013\\
    & PPFD0.8 & 0.014 & 50.411\\
    & PPFD1.2 & 0.009 & 75.616\\
    \bottomhline
    \end{tabular}
\end{table}

@@

The accumulation of ozone dose is strongly depended on the timing and length of the GS. \citet{ICP:MappingManual2017} suggest a simple latitude model for beech, birch, and Norway spruce to localize start and end of the GS:
%
\begin{align}
  A_\mathrm{start} &= 105 + (LAT-50\,\unit{^\circ N})\cdot 1.5\\
  A_\mathrm{end} &= 297 - (LAT-50\,\unit{^\circ N})\cdot 2
\end{align}
%
At Svanhovd, this yields $A_\mathrm{start}=135~\unit{doy}$ and $A_\mathrm{end}=257~\unit{doy}$. We verified these as follows.

For coniferous trees, we used a MODIS (Aqua/Terra) retrieved GPP product \citep{MODIS_PSN} on a $1\times 1\,\unit{km}$ patch centered at Svanhovd. GPP over time follows approximately a second-order polynomial function (Fig.~\ref{fig:modis_gpp}). By calculating the roots of the fitted polynomial, we found $A_\text{start}=122~\unit{doy}$ in 2018 and $A_\text{start}=106~\unit{doy}$ in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. These $A_\mathrm{end}$ values will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 was one month longer than in 2018. For $A_\text{start}$ of deciduous trees, we used temperature-degree-days (above $5\,\unit{^\circ C}$) on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}) which corresponds to the phenological timing of budburst for birch trees. We found \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR) and are very similar to the latitude model predictions. For perennial grassland, we assume a latency period of $1\,\unit{month}$ after snow-melt. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}. Assuming the linear trend for temperture-degree-days defined GS over the past 6~decades prevails, GS at Svanhovd could converge with GS in Central Europe. All results are tabulated in Table~\ref{tab:sensitivity_tests_gs}.

@@

\subsection{Climate and weather conditions in the growing season 2018/19}
\label{sec:stats}
In the following section, we consider the climate local to the Svanhovd region to understand the characteristics that might influence the extent of stomatal ozone dose and hence risk. We conduct this analysis for the two years 2018 and 2019 and relate these years to seasonality to understand how risk is influence by climate.

@@

Climatologies (typically multi-year monthly averages) of temperature, precipitation, and radiant energy $Q_\mathrm{e}$ for Svanhovd were derived for the period 1992--2012 and are shown as box plots in Fig.~\ref{fig:climatediagram}. Therein, the colored horizontal lines indicate the median and the triangles the mean value for the respective month. A box denotes the interquartile range (IQR). Observations exceeding 1.5~times the IQR (whiskers) are usually considered outliers but ought to be interpreted differently in a climate diagram, namely as events deviating extraordinarily from the norm.

>>>>

In Section~\ref{sec:dam}, we give a brief summary of the key components and meteorological input data of the $\mathrm{DO_3SE}$ model that we use to assess the ozone risk. Subsequently, we propose a local climate derived temperature response function and explore the impact of the stomata aperture parameterization at low light intensities. Start and end of the growing season are inferred from satellite observed gross primary productivity (GPP). We explore how risk might be over- or underestimated using Boreal parameterization (Section~\ref{sec:results}). In Section~\ref{subsec:res_local}, we evaluate the local climate derived parameterizations by means of average relative stomatal conductance. For evaluating the $\mathrm{DO_3SE}$ model results in a climate perspective, we assess the climate anomalies of weather conditions and ozone concentrations in the 2018 and 2019 GS at our research site (Section~\ref{subsec:stats}). In Section~\ref{subsec:do3se_results}, we assess the difference in predicted risk for ozone-induced damage for the local climate derived parameterizations compared to MM parameters. We discuss our results concerning prevailing conditions and give an outlook on how future assessments of pollution risk in the subarctic could be improved in Section~\ref{sec:conc}.

%%
\section{Methods}
\label{sec:dam}
In the following, we briefly describe the $\mathrm{DO_3SE}$ model used for assessing the ozone risk, its key components, input data, and gap filling procedure (Section~\ref{sec:do3se}). To study the impact of model parameters, we propose a method to define a parameterization for a subarctic climate based on existing Boreal parameterizations and local climate conditions, study the impact of stomata aperture parameterization at low light intensities, and estimate the start and end of the growing season from satellite derived GPP (Section~\ref{sec:localization}).

@@

The aperture of stomata referred to as $g_\mathrm{sto}$ is governed by different species-specific factors and responds to key environmental variables. In the $\mathrm{DO_3SE}$ model, $g_\mathrm{sto}$ is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g_\text{sto} = g_\text{max} \cdot f_\text{phen} \cdot f_\text{light} \cdot \max{\left\{f_\text{min}, f_\mathrm{T} \cdot f_\mathrm{VPD} \cdot f_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is modified by factors that vary within a range of $0-1$. These factors are empirically determined and account for leaf phenology $f_\text{phen}$, light $f_\text{light}$, temperature $f_\mathrm{T}$, water vapor pressure deficit $f_\mathrm{VPD}$, and soil water potential $f_\text{SWP}$. All factors differ with plant functional type (PFT).

The leaf phenology factor $f_\mathrm{phen}$ is a stepwise defined linear function that allows for an increase of stomatal conductance in spring (start of growing season) and a decrease in fall (end of growing season).

@@

$\Phi_\mathrm{sto}$ for a sunlit upper canopy leaf (vegetation height $h$) is then calculated under the assumption that the $\chem{[O_3]}_h$ at the top of the canopy $h$ represents a reasonable estimate of the \chem{[O_3]} at the upper surface of the laminar layer:
%
\begin{equation}
  \Phi_\mathrm{sto}(t) = \chem{[O_3]}_h(t) \cdot \frac{g_\mathrm{sto}}{g_\mathrm{sto}+g_\mathrm{ext}} \cdot \frac{1}{r_b+r_c}.
  \label{eq:flux_stomata}
\end{equation}
$\chem{[O_3]}_h$ is time dependent and typically 1-hourly averaged. Using this $\mathrm{POD_y}$ is then computed according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}(t)-y)\cdot \mathrm{d}t},
  \label{eq:pod}
\end{equation}
%
with a stomatal ozone flux threshold $\mathrm{y}$ and $\Phi_\mathrm{sto}(t)$ given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}^{-1}\,s^{-1}}. Herein PLA stands for Projected~Leaf~Area. The integration timestep $\mathrm{d}t$ is usually 1~hour. The flux threshold~$\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). Finally, $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

@@

To fill gaps in the ozone data, we applied the methodology described in \citet{ACP:Falk2021} that is based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology (multiannual norm) \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. \citet{ACP:Falk2021} computed a generalized climatology for northern Fennoscandia from long-term records of different close by atmospheric monitoring sites and determined the station with the highest correlation coefficient compared to the site with the gap. For this site, the anomalies with respect to the generalized climatology were calculated for the gap period. These anomalies were than added to the climatology of the site with the gap and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details and a comparison with regional air quality models refer to \citet{ACP:Falk2021}.

@@

For our study focuses on ozone impact on subarctic vegetation, we chose Svanhovd in northern Norway ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$) as site. Reasons for our choice were: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution monitoring (\chem{NO_x}, \chem{SO_x}, particular matter (PM), radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018 and 2019 growing seasons.

All essential climate variables for the 2018 and 2019 growing seasons were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The \chem{O_3} monitoring data have been added to a database for air-quality monitoring called EBAS operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration, drifts in the observed data occurred imposing false trends which are small but in the same order of magnitude as the expected trends in tropospheric background ozone. This does not concern the ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible injuries on ozone sensitive plants, following \citet{GCB:Mills2011}, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd during the 2018 and 2019 GSs. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we observed visible ozone injuries on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible injuries were found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). Note that the extent of visible injuries is, in general, not equivalent to a quantitative reduction of above/below ground biomass of a species.

@@

We determined the dominant species on-site at Svanhovd and from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from Boreal paramterizations when available \citep{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations. In particular, the subarctic local climate derived parameterization for coniferous trees is based on MM's Boreal Norway spruce, deciduous trees on the MM's Boreal silver birch, and perennial grassland on MM's perennial grassland for Central Europe as a Boreal parameterization is not available. For a comprehensive list of MM parameters, consult Supplement~Table~S1.

@@

Initial simulations with the $\mathrm{DO_3SE}$ model using the Central European MM parameterization for perennial grassland revealed an unexpectedly low leaf level stomatal conductance ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occurred only during an extended warm period in late July. In ecological terms, this means that there would have been almost no growth of grass in the summer of 2019 -- a model prediction that is falsified by reality. We identified $f_\mathrm{T}$ as abnormally low throughout the season, being the limiting factor of stomatal conductance in perennial grassland. This suggested that deriving a local climate parameterization is necessary for appropriate modeling.

To derive a subarctic parameterization for the temperature response function $f_\mathrm{T}$, we divide the temperature dataset into two subsets: 1992--2000 (near past) and 2011--2019 (present). From these two subsets, we select only dates within the GS (May--September) and plott the probability density functions (PDFs) as histograms with $1\,\unit{^\circ C}$ binning (Fig.~\ref{fig:f_temp_grassland}a). A comparison between the two PDFs indicates that during the past decade the temperature distribution has shifted towards higher temperatures reducing the number of days below $5\,\unit{^\circ C}$. It is unclear how fast the species composition or physiology of perennial grassland responds to these changes. Thus, the predominant acclimation at present is not known. Further, we overlay these histograms with the MM $f_\mathrm{T}$. A low overlapping area suggests a low agreement with local climate conditions.

We define local climate derived parameterizations for two hypothetical species (\emph{subarctic} and \emph{cold}) by increasing the overlapping area between $f_\mathrm{T}$ and the temperature PDF during the GS. Then we construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures than the MM species. This is accomplished by moving $T_\mathrm{min}$ and $T_\mathrm{opt}$ towards cooler temperatures but $T_\mathrm{max}$ is kept at its MM value. In the same way, \emph{subarctic} is constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift both $T_\mathrm{min}$ and $T_\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. The most extreme of the two local climate derived parameterizations compared to the MM parameterization is the \emph{subarctic}. This methodology is analogous to the optimal temperature acclimation experimentally found for photosynthesis \citep{PR:Way2013}.

For localizing $f_\mathrm{light}$, we presume that the aperture of stomata at low light conditions differs in subarctic species compared to species in less extreme climates. We, therefore, adjust the modelled light sensitivity of stomata such that the PPFD value needed to reach $50\,\unit{\%}$ opening is varied by $\pm 20\,\unit{\%}$. We derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT $k$ analytically
%
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
%
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value of $\alpha_\mathrm{MM}^k$ at $0.5$ ($50\,\%$ opening)
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. By solving Eq.~(\ref{eq:inverse_function}) for $\alpha$, we find $\alpha'(\gamma')$.

In Fig.~\ref{fig:f_temp_grassland}b, we show the PPFD PDF for near past and present together with the $f_\mathrm{light}$ for perennial grassland. The Central European MM parameterization is displayed as solid line for comparison.

@@

Similarly, we define local climate derived $f_\mathrm{T}$ and localized $f_\mathrm{light}$ for coniferous and deciduous trees. The resulting response functions are shown in Appendix~Figs.~\ref{fig:f_temp_spruce} and \ref{fig:f_temp_birch}, respectively.
Coniferous trees are known to start photosynthesis as soon as the leaf temperature reaches a certain threshold \citep{TB:Kolari2007}. Thus, coniferous trees are already active at low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as day of the year~(\unit{doy})~100 but cease activity in less favorable conditions \citep{TB:Kolari2007, TP:Wallin2013}. 
To determine the local climate derived temperature parameterization for coniferous trees, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007}. From these, we estimate an optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). For simplicity, the same interval is assumed for deciduous trees.

@@

\begin{table}[t]
  \caption{Localized light response parameterizations. MM refers to mapping manual \citep{GCB:Mills2011,ICP:MappingManual2017}.}
  \label{tab:sensitivity_tests_light}
  \begin{tabular}{llcr}
    \tophline
    Species & type & $\alpha$ & $\gamma(0.5)$\\
    \middlehline
    \multirow{3}{*}{Deciduous tree} & MM & 0.004 & 165.035\\
    & PPFD0.8 & 0.005 & 132.028\\
    & PPFD1.2 & 0.003 & 198.042\\
    \middlehline
    \multirow{3}{*}{Coniferous tree} & MM & 0.006 & 115.525\\
    & PPFD0.8 & 0.008 & 92.420\\
    & PPFD1.2 & 0.005 & 138.629\\
    \middlehline
    \multirow{3}{*}{Perennial grassland} & MM & 0.011 & 63.013\\
    & PPFD0.8 & 0.014 & 50.411\\
    & PPFD1.2 & 0.009 & 75.616\\
    \bottomhline
    \end{tabular}
\end{table}

@@

The accumulation of ozone dose is strongly depended on the timing and length of the GS. \citet{ICP:MappingManual2017} suggested a simple latitude model for beech, birch, and Norway spruce to localize start and end of the GS:
%
\begin{align}
  A_\mathrm{start} &= 105 + (LAT-50\,\unit{^\circ N})\cdot 1.5\\
  A_\mathrm{end} &= 297 - (LAT-50\,\unit{^\circ N})\cdot 2
\end{align}
%
At Svanhovd, this yields $A_\mathrm{start}=135~\unit{doy}$ and $A_\mathrm{end}=257~\unit{doy}$. We verify these as follows.

For coniferous trees, we use a MODIS (Aqua/Terra) retrieved GPP product \citep{MODIS_PSN} on a $1\times 1\,\unit{km}$ patch centered at Svanhovd. GPP over time follows approximately a second-order polynomial function (Fig.~\ref{fig:modis_gpp}). By calculating the roots of the fitted polynomial, we find $A_\text{start}=122~\unit{doy}$ in 2018 and $A_\text{start}=106~\unit{doy}$ in 2019. $A_\text{end}$ amounts to \unit{doy}~261 and 274, respectively. These $A_\mathrm{end}$ values are used for all PFTs alike. The resulting growing season for coniferous trees in 2019 is one month longer than in 2018. For $A_\text{start}$ of deciduous trees, we use temperature-degree-days (above $5\,\unit{^\circ C}$) on gridded temperature data from SeNorge.no (Appendix~Fig.~\ref{fig:greening_season_change_Svanvik}) which corresponds to the phenological timing of budburst of birch trees. We find \unit{doy}~129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the closest inland meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR) and are very similar to the latitude model predictions. For perennial grassland, we assume a latency period of $1\,\unit{month}$ after snow-melt. This assumption is supported by observations in Rovaniemi, Finland \citep[][Supplement~Fig.~S1]{FCR:Korhonen2018}. Supposed, the linear trend for temperture-degree-days defined GS over the past 6~decades prevails, GS at Svanhovd could converge with GS in Central Europe by the end of the century. All GS results are tabulated in Table~\ref{tab:sensitivity_tests_gs}.

@@

\subsection{Climate and weather conditions in the growing seasons 2018 and 2019}
\label{subsec:stats}
In the following section, we consider the climate local to the Svanhovd region to understand the characteristics that might influence the extent of stomatal ozone dose and hence risk. We conduct this analysis for the two years 2018 and 2019 and relate these years to seasonality to understand how risk for ozone-induced damage is influence by climate.

@@

Climatologies (typically multi-annual monthly averages) of temperature, precipitation, and radiant energy $Q_\mathrm{e}$ for Svanhovd were derived for the period 1992--2012 and are shown as box plots in Fig.~\ref{fig:climatediagram}. Radiant energy is the diurnal integral of $Q_0$. The colored horizontal lines indicate the median and the triangles the mean value for the respective month. A box denotes the interquartile range (IQR). Observations exceeding 1.5~times the IQR (whiskers) are usually considered outliers but ought to be interpreted differently in a climate diagram, namely as events deviating extraordinarily from the norm.
